Analyzing Student Interaction Patterns with GenAI-Powered Chatbots in Programming Courses through Metacognitive Strategies of Spacing and Interleaving

### 2024-11-19 AEID2025/PaperAEID25.tex
PaperAEID25

\documentclass[a4paper,twoside]{article}

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
% Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\usepackage{natbib}
\usepackage{multirow}
\usepackage{SCITEPRESS}
\usepackage{arydshln}
\widowpenalty=10000
\clubpenalty=10000

\begin{document}

\title{
  Analyzing Student Use of Spacing and Interleaving Strategies in Interactions
  with GenAI-Powered Chatbots in Programming Courses
}

\author{\authorname{Rodrigo Prestes Machado\sup{1}\orcidAuthor{0000-0003-0428-6387},
Carlos Alario-Hoyos\sup{2}\orcidAuthor{0000-0002-3082-0814},
Patricia Callejo\sup{2}\orcidAuthor{0000-0001-6124-6213},
Iria Estévez-Ayres\sup{2}\orcidAuthor{0000-0002-1047-5398},
Carlos Delgado Kloos\sup{2}\orcidAuthor{0000-0003-4093-3705}}
\affiliation{\sup{1}Department of Informatics, Instituto Federal de Educação,
Ciência e Tecnologia, Porto Alegre, Brazil}
\affiliation{\sup{2}Department of Telematics Engineering, Universidad Carlos III
de Madrid, Leganés (Madrid), Spain}
\email{\ rodrigo.prestes@poa.ifrs.edu.br, \{calario, pcallejo, ayres, cdk\}@it.uc3m.es}
}

\keywords{
  Programming, Generative Artificial Intelligence, Chatbots, Spacing,
  Interleaving and Metacognition
}

\abstract{
The aim of this study is to analyze the interactions of programming students
with a chatbot powered by OpenAI's GPT-3.5, enhanced with the Retrieval
Augmented Generation (RAG) technique, within the context of a Java programming
course. The study investigates how students use the chatbot while employing
two metacognitive strategies: interleaving and spacing. The methodology involved
classifying and categorizing the students' interaction prompts with the chatbot
into eight distinct categories. The findings reveal that, while most
interactions focus on active learning strategies, students demonstrate limited
conscious application of metacognitive strategies. The study highlights the need
for more comprehensive guidance on the strategic use of AI tools to enhance
learning outcomes.
}

\twocolumn\maketitle\normalsize\setcounter{footnote}{0}

\section{\uppercase{Introduction}}
\label{sec:introduction}

Recent advances in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming \citep{Peng23}
\citep{Pandey24} and education \citep{Puryear22}. In programming courses,
students can employ GenAI tools to improve their understanding, receive
personalized feedback, and access detailed explanations. For example, GitHub
Copilot, a tool that is integrated into development environments, assists in
providing real-time suggestions and accelerating code writing, which could help
streamline the learning process \citep{Denny23}. Educational chatbots, on the
other hand, try to guide students through coding challenges, answer questions,
and try to foster a deeper understanding of programming concepts
\citep{Labadze23}.

The study conducted by \cite{chan23} revealed that both undergraduate and
postgraduate students have positive attitudes towards the use of GenAI.
A systematic review organized by \cite{Lo24} demonstrated that students can
effectively learn from ChatGPT, resulting in improved comprehension and academic
achievement \citep{Callejo24}. Furthermore, it was observed that ChatGPT may
allow students to regulate their learning pace \citep{Baha24} and can
potentially support self-regulated learning, particularly for students with
prior technical and disciplinary knowledge \citep{Xia23}.

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54.5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \citep{Boudouaia24}. All of these studies highlight
the need to increase the understanding of how students interact with these tools
and how they can be used to improve learning, as emphasized by \cite{Lo24}.

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational chatbot powered by OpenAI gpt-3.5, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of two metacognitive strategies:
spacing \citep{Carvalho20} and interleaving \citep{Rivers21}. To achieve this
objective, we formulate three research questions:

@@

\section{\uppercase{Theoretical Framework}}

@@

Given that GenAI often produces highly accurate and automatic responses
\citep{Puryear22}, it is essential that students use these tools within an
active learning process to enhance their learning outcomes. This active
learning process can be further understood through the lens of metacognition,
which focuses on awareness of one's mental processes. \cite{flavell79}
proposed a model of metacognitive monitoring that includes four interrelated
phenomena: Metacognitive Knowledge, Metacognitive Experience, Metacognitive
Goals, and Metacognitive Actions. These processes do not occur in isolation, but
they influence each other, altering cognitive progress over time.

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise before,
during, and after cognitive activity, such as frustration, confusion,
satisfaction, and others. Metacognitive Goals are the key to regulating thought
as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\citep{Ouhao18, Carvalho20}, interleaving topics \citep{Rivers21}, utilizing
retrieval practice \citep{larsen18}, and other possible strategies.

Spacing and interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and support the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\citep{Carvalho20}. Interleaving involves mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \citep{Rivers21}.

@@

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{4cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change (LC) & Prompts that request a change of idiom. \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

\section{\uppercase{Material and Method}}

@@

CharlieBot is an educational chatbot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \citep{Sun24}. The RAG approach was developed
using materials from a MOOC, including text, exercises, video transcripts, and
other resources.

@@

\section{\uppercase{Results and Discussion}}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.62]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.45]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

@@

For the analysis of the spacing, we initially considered a minimum interval of
60 minutes as a reference to define the significant spacing \citep{Gadella24}.
Based on this, our results showed that, of the 81 conversations analyzed, only
27 (33.3\%) had at least one message with an interval greater than 60 minutes.
Figure 2 illustrates the elapsed time between messages (delta). For example,
in conversation 15, the first message was sent at time zero, while the second
message was sent by the student after a delta of 0.6 minutes. Additionally,
message 4 in conversation 15, as shown in Figure \ref{fig:graph2}, presents a
delta of 69.1 minutes, which, according to our reference, characterizes
significant spacing. Figure \ref{fig:graph3} also shows the number of spacings
in each conversation. Among conversations that exhibited spacing, the
average number of subsections per conversation was 2.2. Therefore, a
conversation with one instance of spacing is typically divided into two distinct
study subsections.

@@

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.39]{img/figure3.png}
  \caption{Subsections per conversation. \textit{A zero value indicates
  no subsection and no spacing.}}
  \label{fig:graph3}
\end{figure*}

@@

@@

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{4cm} p{4cm} p{4cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.26 & 1.74 \\
      Uncategorized & 1.65 & 3.54 \\
      Language Change & 1.89 & 3.28 \\
      Multiple Question & 2.65 & 6.10 \\
      Student Correction & 3.17 & 7.33 \\
      \hdashline
      \textit{Overall} & \textit{3.43} & \textit{6.60} \\
      \hdashline
      Debugging Help & 4.84 & 10.13 \\
      Conceptual Questions & 5.41 & 10.54 \\
      Complete Solution & 6.55 & 10.14 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

It is important to consider how GenAI tools impact students in various
demographic groups, academic disciplines, cultural backgrounds, and levels of
previous experience \citep{catalan21} \citep{neo22}. Consequently, this study is
limited to a specific group of students and focuses on the use of a single GenAI
tool. As a result, the findings may not be generalizable to other populations or
tools.

@@

\bibliographystyle{apalike}
{\small
\bibliography{References}}

\end{document}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\usepackage{arydshln} % Add this line to use the \hdashline command

\begin{document}
%
\title{Analyzing Student Interaction Patterns with GenAI-Powered Chatbots in
Programming Courses through Metacognitive Strategies of Spacing and Interleaving
}
%
\titlerunning{Analyzing Student Interaction with Chatbots through Metacognition}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{
Rodrigo Prestes Machado\inst{1}\orcidID{0000-0003-0428-6387} \and
Carlos Alario-Hoyos\inst{2}\orcidID{0000-0002-3082-0814} \and
Patricia Callejo\inst{2}\orcidID{0000-0001-6124-6213} \and
Iria Estévez-Ayres\inst{2}\orcidID{0000-0002-1047-5398} \and
Carlos Delgado Kloos\inst{2}\orcidID{0000-0003-4093-3705}}
%
\authorrunning{Machado, R. P., et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Instituto Federal de Educação, Ciência e Tecnologia, Porto Alegre, Brazil \\
\url{http://poa.ifrs.edu.br} \\
\email{rodrigo.prestes@poa.ifrs.edu.br} \and
Universidad Carlos III de Madrid, Leganés, Spain \\
\url{https://www.uc3m.es} \\
\email{\{calario, pcallejo, ayres, cdk\}@it.uc3m.es}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The aim of this study is to analyze the interactions of programming students
with a chatbot powered by OpenAI's GPT-3.5, enhanced with the Retrieval
Augmented Generation (RAG) technique, within the context of a Java programming
course. The study investigates how students use the chatbot while employing
two metacognitive strategies: interleaving and spacing. The methodology involved
classifying and categorizing the students' interaction prompts with the chatbot
into eight distinct categories. The findings reveal that, while most
interactions focus on active learning strategies, students demonstrate limited
conscious application of metacognitive strategies. The study highlights the need
for more comprehensive guidance on the strategic use of AI tools to enhance
learning outcomes.

\keywords{Programming \and Generative Artificial Intelligence \and Chatbots \and
Spacing \and Interleaving \and Metacognition.}
\end{abstract}
%
%
\section{Introduction}

Recent advances in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming \cite{Peng23}
\cite{Pandey24} and education \cite{Puryear22}. In programming courses,
students can employ GenAI tools to improve their understanding, receive
personalized feedback, and access detailed explanations. For example, GitHub
Copilot, a tool that is integrated into development environments, assists in
providing real-time suggestions and accelerating code writing, which could help
streamline the learning process \cite{Denny23}. Educational chatbots, on the
other hand, try to guide students through coding challenges, answer questions,
and try to foster a deeper understanding of programming concepts
\cite{Labadze23}.

The study conducted by \cite{chan23} revealed that both undergraduate and
postgraduate students have positive attitudes towards the use of GenAI.
A systematic review organized by \cite{Lo24} demonstrated that students can
effectively learn from ChatGPT, resulting in improved comprehension and academic
achievement \cite{Callejo24}. Furthermore, it was observed that ChatGPT may
allow students to regulate their learning pace \cite{Baha24} and can
potentially support self-regulated learning, particularly for students with
prior technical and disciplinary knowledge \cite{Xia23}.

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54.5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \cite{Boudouaia24}. All of these studies highlight
the need to increase the understanding of how students interact with these tools
and how they can be used to improve learning, as emphasized by \cite{Lo24}.

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational chatbot powered by OpenAI gpt-3.5, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of two metacognitive strategies:
spacing \cite{Carvalho20} and interleaving \cite{Rivers21}.

To achieve this
objective, we formulate three research questions:

@@

\section{Theoretical Framework}

@@

Given that GenAI often produces highly accurate and automatic responses
\cite{Puryear22}, it is essential that students use these tools within an
active learning process to enhance their learning outcomes. This active
learning process can be further understood through the lens of metacognition,
which focuses on awareness of one's mental processes. \cite{flavell79}
proposed a model of metacognitive monitoring that includes four interrelated
phenomena: Metacognitive Knowledge, Metacognitive Experience, Metacognitive
Goals, and Metacognitive Actions. These processes do not occur in isolation, but
they influence each other, altering cognitive progress over time.

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise before,
during, and after cognitive activity, such as frustration, confusion,
satisfaction, and others. Metacognitive Goals are the key to regulating thought
as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\cite{Ouhao18,Carvalho20} interleaving topics \cite{Rivers21}, utilizing
retrieval practice \cite{larsen18}, and other possible strategies.

Spacing and interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and support the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\cite{Carvalho20}. Interleaving involves mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \cite{Rivers21}.

@@

@@

\section{Material and Method}

@@

CharlieBot is an educational chatbot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \cite{Sun24}. The RAG approach was developed
using materials from a MOOC, including text, exercises, video transcripts, and
other resources.

@@

\begin{table}[htbp]
\caption{Categories - adapted from \cite{Ghimire24}}
\centering
\renewcommand{\arraystretch}{1.2} % Increase the spacing between rows
\begin{tabular}{|p{4cm}|p{8cm}|} % Add vertical bars
\hline
\textbf{Category} & \textbf{Description} \\
\hline
Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. \\
Student Correction (SC) & Prompts where the student corrects the chatbot. \\
Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. \\
Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. \\
Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
Language Change (LC) & Prompts that request a change of idiom. \\
Uncategorized (U) & Prompts that do not fit into any of the above categories. \\
\hline
\end{tabular}
\label{tab:categories}
\end{table}

\section{Results and Discussion}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.8]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

@@

For the analysis of the spacing, we initially considered a minimum interval of
60 minutes as a reference to define the significant spacing \cite{Gadella24}.
Based on this, our results showed that, of the 81 conversations analyzed, only
27 (33.3\%) had at least one message with an interval greater than 60 minutes.
Figure 2 illustrates the elapsed time between messages (delta). For example,
in conversation 15, the first message was sent at time zero, while the second
message was sent by the student after a delta of 0.6 minutes. Additionally,
message 4 in conversation 15, as shown in Figure \ref{fig:graph2}, presents a
delta of 69.1 minutes, which, according to our reference, characterizes
significant spacing. Figure \ref{fig:graph3} also shows the number of spacings
in each conversation. Among conversations that exhibited spacing, the
average number of subsections per conversation was 2.2. Therefore, a
conversation with one instance of spacing is typically divided into two distinct
study subsections.

\begin{figure*}[htbp]
    \centering
    \includegraphics[scale=0.35]{img/figure2.png}
    \caption{Examples of conversations.}
    \label{fig:graph2}
\end{figure*}

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.3]{img/figure3.png}
  \caption{Subsections per conversation. \textit{A zero value indicates no
  subsection and no spacing.}}
  \label{fig:graph3}
\end{figure*}

@@

@@

\begin{table}[htbp]
\caption{Average and Standard Deviation of Different Categories (in minutes)}
\centering
\renewcommand{\arraystretch}{1.2} % Increase the spacing between rows
    \begin{tabular}{|p{4cm}|p{4cm}|p{4cm}|} % Add vertical bars
        \hline
        \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
        \hline
        Code Snippet & 1.26 & 1.74 \\
        Uncategorized & 1.65 & 3.54 \\
        Language Change & 1.89 & 3.28 \\
        Multiple Question & 2.65 & 6.10 \\
        Student Correction & 3.17 & 7.33 \\
        \hdashline
        \textit{Overall} & \textit{3.43} & \textit{6.60} \\
        \hdashline
        Debugging Help & 4.84 & 10.13 \\
        Conceptual Questions & 5.41 & 10.54 \\
        Complete Solution & 6.55 & 10.14 \\
        \hline
    \end{tabular}
\label{tab:averages}
\end{table}

@@

@@

It is important to consider how GenAI tools impact students in various
demographic groups, academic disciplines, cultural backgrounds, and levels of
previous experience \cite{catalan21} \cite{neo22}. Consequently, this study is
limited to a specific group of students and focuses on the use of a single GenAI
tool. As a result, the findings may not be generalizable to other populations or
tools.

@@

\begin{credits}
\subsubsection{\ackname}
Thanks to Instituto Federal de Educação, Ciência e Tecnologia do Rio Grande do
Sul (IFRS) for partially funding this research
\end{credits}
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{References}
\end{document}
